---
title: "RLadies ggplot2"
author: "Sarah Hosking"
date: "March 8, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lubridate)
library(tidyverse)
library(GGally)

library(reshape2) #plot missing vals

theme_set(theme_bw())

```

# Why ggplot?

Big advantages:

* consistent structure regardless of plot type
* easy to iterate
* *excellent grouping and faceting*
* requires tidy data
* plots look pretty good OOTB (albeit with room for improvement)


Some complaints:

* sometimes can be verbose
* slow?
* not as good for print publication - ?


# Paris Air Pollution

Because we all need a break from car, iris and most of all, election, data.

More seriously, it was because last December I found myself repeatedly wondering:

* So just how bad is Paris' air pollution, really?
and
* Is there a bad time to go running?

## Load Data

```{r}

airparif <- read_rds('airparif.rds')
dim(airparif)
```

```{r}
str(airparif)
```

```{r}
head(airparif)
```

```{r}
tail(airparif)
```

```{r}
summary(airparif)
```


## What are we looking at?


```{r}

names(airparif)

```

What the abbreviations mean:

* `PM25` = fine particulate < 2.5 mm
* `PM10` = fine particulate < 10 mm
* `03` = Ozone
* `NO2` = Nitrogen Dioxide (Azote)
* `CO` = Carbon monoxide


# Explore distributions

### DEMO: Histogram

```{r}

ggplot(data = airparif, aes(x = PM10)) +
  geom_histogram(binwidth = 5)

```


You can also use expressions for your data
```{r}
ggplot(data = airparif, aes(x = PM10 + PM25)) +
  geom_histogram(binwidth = 5)

```

### HANDS ON: Examine the other pollutants

Explore `PM25`, `NO2`, `O3`, `CO`. Anything stand out?

```{r explore distributions}

# basic ggplot template for univariate plots

p <- ggplot(airparif, aes(x=NO2)) +
  geom_histogram()


```

```{r add vert line}

# "low" levels are > 25. Mark with a vertical line
p + 
  geom_vline(xintercept = 25, colour = "lightgreen")

```

### DEMO: How does the distribution change over the year?

To find out let's facet by month.

```{r}

p +
  geom_vline(xintercept = 25, colour = 'lightgreen') + 
  facet_wrap(~month)

```


### DEMO: How does the distribution change over the years?

To find out let's facet by month

```{r}

p +
  geom_vline(xintercep=t = 25, colour = 'lightgreen') + 
  facet_wrap(~month)

```

### HANDS ON: What do the boxplots look like? 

```{r}

p <- ggplot(airparif, aes(year, NO2)) 

p + 
  geom_boxplot()
```

#### Including scales

To zoom in, you can add limits to the x or y scale. But there's a catch

```{r avoid this}
 p +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 60))

```

Notice how the proportions and positions of the boxes have changed. That's because we've removed all the points above 60.

Instead, use `coord_cartesian`. This creates a true zoom, without distorting IQR.

```{r do this}

p + 
  geom_boxplot() + 
  coord_cartesian(ylim = c(0,60))

```


# Adding more variables to plots

To segment your plot into different groups, you can specify a variable for color, shape, linetype etc.

### DEMO: Add a grouping variable

What else happens when you include a "grouping"" variable in the aesthetic layer?

```{r easy iteration + grouping}
p <- ggplot(data = airparif, aes(x = PM25))

# make a frequency polygon
p + geom_freqpoly()

# change the colour
p + geom_freqpoly(color = 'blue')


```

### HANDS ON: show frequency polygons for each year in the same plot.

Hint: use aesthetics
The aesthetics available for geom_freqpoly are:

* alpha

* colour

* group

* linetype

* size

```{r hands on}

# group by year
# what gets added to the chart?
p <- ggplot(airparif, aes(PM25))

p <- p +
  geom_freqpoly(aes(colour = year))


# or group by month

```

With this type of plot, facets would be better.

### HANDS ON: Facet the previous plot by month.

This will create a grid of plots, one for each month.

```{r facet previous plot by year}
p +
  facet_wrap(~year)

```

# Look at correlations

### DEMO: Plot a correlation matrix

```{r ggpairs, eval=FALSE, warning=FALSE, include=FALSE}

# create a sample - ggpairs takes time.
set.seed(888)

# limit the number of columns
sample <- subset(airparif, select = c(PM10, PM25, NO2, O3, CO, month))
# limit the number of rows
sample <- sample[sample(1:nrow(sample),1000),]

# create simple correlation matrix plot
ggpairs(data=sample,
        columns=1:3)

# more complex
ggpairs(data=sample,
        #columns=1:3
        title="Pollution Correlation Matrix")


# add smoothing & modify scatterplots, fix bins
ggpairs(data=sample,
        #columns=1:3
       lower = list(continuous = wrap("smooth", alpha=1/5, shape = I('.')),
                     combo = wrap("facethist", binwidth = 5)), 
      title="Pollution Correlation Matrix")
```

# Summarize and plot

How polluted is Paris? How many days of higher levels of pollution were there?

*To answer this we first need to tidy the data.*

### DEMO: Gather pollutants into one column

```{r}

# Option 1: load the summarized data
airparif.long <- read_rds("airparif_long.rds")


# Option 2: summarize data yourself

airparif.long <- airparif %>% 
   gather(PM10,PM25,NO2,O3,CO, 
         key = 'pollutant',
         value = 'value', 
         na.rm = TRUE) %>% 
   group_by(date, pollutant) %>%
   summarise('mean' = mean(value),
             'median' = median(value),
             'max' = max(value))

head(airparif.long)

```

Now plot this new df

```{r}

ggplot(aes(date, median, colour = pollutant), 
       data = airparif.long) +
         geom_line()

```

Wow, carbon monoxide (`CO`) looms over everything else.

### DEMO: Drill down on fine particulate

```{r}

# only plot PM pollutants
particles <- airparif.long %>% 
  filter(pollutant %in% c('PM10', 'PM25')) 


p <- ggplot(particles, aes(date, median))


p + 
  geom_line(aes(colour = pollutant)) +
  scale_x_date(date_breaks = '3 months', date_labels = "%b %y")
  
```

Need to add some more context to this plot. What are the levels for medium, high, and very high pollution?


## How many days at each pollution level?

** 

```{r}

# Option 1: load the summarized data
#levels <- read_rds("levels.rds")


# Option 2: summarize data yourself

# First, load function to tag pollution levels

# OLD FUNCTION - IGNORE
# credit to https://www.r-bloggers.com/r-function-of-the-day-cut-2/

# pollution.level <- function(x, lower = 0, upper = 100, 
#                             by = 25, sep = "-", above.char = "+") {
 
# labs <- c("very low", "low", "medium", "high", "very high")
# 
# cut(floor(x), 
#     breaks = c(seq(lower, upper, by = by), Inf),
#     right = FALSE, 
#     labels = labs)

# USE THIS INSTEAD
pollution.level <- function(x) {
 
 labs <- c("very low", "low", "medium", "high", "very high")

 cut(x, 
     breaks = c(0,15,30,50,100,999),  # NOTE: in reality, only applies to PM10!!
     right = TRUE,
     #include.lowest = TRUE, #new setting!
     labels = labs)
}

# Then complete this code
levels <-  airparif.long %>%
  group_by(date, pollutant) %>% 
  mutate_all(funs(pollution.level))

```


### DEMO: Plot days per level per pollutant

How many days at each level for each pollutant?

```{r}

# create a bar plot
p <- ggplot(levels, aes(median, fill = pollutant))
p + geom_bar()

```

Make this easier to read

```{r}
p + 
  geom_bar(position = 'dodge')

```

Still hard to compare the distribution of levels for each pollutant. 

### HANDS ON: How would you improve this?

We can facet by pollutant.

```{r}
p + 
  geom_bar(position = 'dodge') +
  facet_wrap(~pollutant)

```

### DEMO: Does this change much from year to year? 

We'll need to add a year column.
We can do this again with dplyr and include ggplot using the pipe

```{r add year var to levels df}

levels %>% 
  # separate(date, c("year", "month", "day"), sep="-") %>% 
  mutate('year' = year(date)) %>% 
  ggplot(aes(median, fill = pollutant)) + 
  geom_bar(position = 'dodge') +
  facet_grid(pollutant ~ year, scales = 'free_y')

```

It doesn't look like it. But if we allow the y-scale to vary, it might give us more insight.


### HANDS ON: What about by month?

You'll need to add month, similar to adding year above. Dplyr has a function to make this easy.

BONUS: Look up dplyr's `separate()` function in the help.

```{r }

levels %>% 
  separate(date, c("year", "month", "day"), sep="-") %>% 
  ggplot(aes(median, fill = pollutant)) + 
  geom_bar(position = 'dodge') +
  facet_grid(pollutant ~ month)

```

Hard to tell, let's free up the y scale for each pollutant

### HANDS ON: Make the y-axis scale flexible

HINT: Look up `facet_grid` in the help, and look for `scales`.

```{r}

levels %>% 
  separate(date, c("year", "month", "day"), sep="-") %>% 
  ggplot(aes(median, fill = pollutant)) + 
  geom_bar(position = 'dodge') +
  facet_grid(pollutant ~ month, scales = "free_y")

```


### DEMO: Make plot more readable

Lastly, let's fix those x-axis labels

```{r fix axis labels}

p <- levels %>% 
  separate(date, c("year", "month", "day"), sep="-") %>% 
  ggplot(aes(median, fill = pollutant)) + 
  geom_bar(position = 'dodge') +
  facet_grid(pollutant ~ month, scales = "free_y")

p +
  scale_y_continuous(breaks = seq(0,150, 50)) +
  scale_x_discrete(name = "Pollution Level") +
  theme(axis.text.x = element_text(size = 4, angle = -45, hjust = -0.1))
  

```



# BONUS STUFF (If we have time)

## Make plots prettier

```{r}

# only plot PM pollutants
# particles <- airparif.long %>% 
#   filter(pollutant %in% c('PM10', 'PM25')) 
# 
# 
# p <- ggplot(particles, aes(date, mean))
# 
# p + 
#   geom_line(aes(y = median, colour = pollutant)) + 
#   scale_x_date() + 
#   theme_
  


```


## qplot

`qplot` is part of `ggplot2`, and stands for "quick plots". Its syntax is similar to base plotting, and is less verbose than `ggplot()`.

```{r qplot hists}

qplot(PM10, data = airparif)

# increase bins
qplot(data = airparif, PM10, binwidth = 5)

```


It's meant, however, to work with vectors. Here's a demo.

```{r}

# qplot meant to work with vectors
a <- c('A', 'B', 'C')
b <- c(1, 2, 3)

# qplot
q <- qplot(a,b)
q

```


```{r}

# ggplot needs a dataframe
d <- data.frame(a,b)

# ggplot
p <- ggplot(data = d, aes(a,b)) +
  geom_point()
p


```


```{r}

# change var
a <- c('X','Y','Z')

# with qplot
q
```

```{r}

# with ggplot
p
```

## Visualize missing values with ggplot

```{r}
# credit to:
# http://www.njtierney.com/r/missing%20data/rbloggers/2015/12/01/ggplot-missing-data/
missing.data <- function(x){
  
  x %>% 
    is.na %>%
    melt %>%
    ggplot(data = .,
           aes(x = Var2,
               y = Var1)) +
    geom_raster(aes(fill = value)) +
    scale_fill_grey(name = "",
                    labels = c("Present","Missing")) +
    theme_minimal() + 
    theme(axis.text.x  = element_text(angle=45, vjust=0.5)) + 
    labs(x = "Variables in Dataset",
         y = "Rows / observations")
}

#missing.data(airparif[,1:7])
```

# Takeaways

Compare ggplot2 to R base plots 
https://flowingdata.com/2016/03/22/comparing-ggplot2-and-r-base-graphics/
http://simplystatistics.org/2016/02/11/why-i-dont-use-ggplot2/
http://varianceexplained.org/r/why-I-use-ggplot2/

qplot or ggplot?
http://stackoverflow.com/questions/5322836/choosing-between-qplot-and-ggplot-in-ggplot2

grouping vs aesthetics
http://stackoverflow.com/questions/10357768/plotting-lines-and-the-group-aesthetic-in-ggplot2

Ggplot cheatsheet: https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf
